---
title: "Exploratory Report on the Weights of Juvenile Snowshoe Hares"
author: "Meghna Rao"
date: "11/7/2021"
output: html_document
code_folding: hide
---
# Overview {.tabset .tabset-fade}
### Introduction 

This report provides an exploratory analysis on data on juvenile snowshoe hares collected from the Bonanza Creek Experimental Forest in Alaska (Kielland, K.et al. (2017)). Juvenile snowshoe hares were trapped and counted over 13 years (1998 - 2012) from three locations: Bonrip, Bonanza Mature,and Bonanza Riparian. Finally, the report analyzes relationships between hindfoot length (mm) and weights (g).

**Data Citation:** [Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative](https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14)

### Data and Analysis

Snowshoe hare trap count, weight and hind foot length data were
collected and made available by Senior Investigator Knut Kielland and
colleagues at the Bonanza Creek (LTER) site in Alaska. Data was collected over the course of 13 years on 378 juvenile hares. Following exploratory data visualization, juvenile hare male and female weights are compared by two-sample t-tests using a significance
level (α) f 0.05 throughout. Differences between the groups are described by Cohen's *d* effect size. The relationship between weight and hind foot length is explored by a simple linear regression. All analyses are in R version 4.0.2 using R studio version 1.2.1056.

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(lubridate) # mutate date columns
library(kableExtra) # to show table statistics
library(effsize) # effect size
library(broom) 
library(ggbeeswarm)
```

```{r}
bonanza_hares <- read.csv(here("data", "bonanza_hares.csv"))

juvenile_hares <- bonanza_hares %>% 
  filter(age == "j") %>%   # Filter for just the juveniles
  mutate(date_new = mdy(date)) %>% # used lubridate library package to create a new column for the date
  mutate(year = year(date_new)) %>% 
   mutate(sex = case_when( # changing names of sex
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",)) %>% 
  mutate(grid = case_when(
    grid == "bonbs" ~ "Bonanza Black Spruce", 
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian")) # change names so it comes up nicely in the visual# created a new column of just the years
   
```

```{r}
juvenile_hares_count <- juvenile_hares %>% 
  count(year) # counted the number of juvenile bunnies per year

juvenile_hares_summary <- juvenile_hares_count %>% 
  summarize(mean_count = mean(n), # mean
            median_count = median(n), # median
            max_count = max(n), # max
            min_count = min(n)) # min
```
## Preliminary Analysis on Annual Juvenile Hare Trap Counts

```{r, fig.cap= "Figure 1: Graph showing the total number of juvenile hares trapped between 1999 and 2012 in the Bonanza Creek Experimental Forest, Tanana Valley, Alaska. Data: Kielland et al. (2017)."}
ggplot(data = juvenile_hares_count, aes(x = year, y = n)) +
  geom_line(color = "mediumblue") +
  labs(x = "Year", y = "Total Juvenile Hares Trapped", title = "Figure 1: Total Juvenile Hares Trapped by Year from 1998 to 2012") + # add labels to graph
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100, 125)) + # manually putting tics for y axis
  scale_x_continuous(breaks = c(1998, 1999, 2000,2001, 2002, 2003, 2004, 2005, 
                                2006, 2007, 2008, 2009, 2010, 2011,2012)) + # manually putting ticks for x axis
  theme_minimal() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5)) # center the title and bold it
```
The average number of juvenile hares trapped per year in this study was `r round(juvenile_hares_summary$mean_count, 0)` and the median number of juvenile hares trapped per year was `r round(juvenile_hares_summary$median_count, 0)`. The greatest number of juvenile hares trapped was `r round(juvenile_hares_summary$max_count, 0)` in the first year of the study in 1999. There were no juvenile hares trapped in 2002 and 2009. The least number of juvenile hares trapped was `r round(juvenile_hares_summary$min_count, 0)` in 2012. To standardize the juvenile hare population in future analyses, the number of juvenile hare traps laid out every year, as well as the time of year that the traps are laid out, need to be consistent from year to year.

## Visualizing Juvenile Hare Weights by Sex and Location

```{r}
juvenile_hares_sex_site <- juvenile_hares %>% 
  group_by(sex, grid) %>% # group by sex, and grid
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```


```{r, fig.cap = "Figure 2: Graph showing summary statistics on the average weight of male and female juvenile hares trapped between 1999 and 2012 in mature lowland black spruce (Bonanza Black Spruce), mature black spruce forest (Bonanza Mature) and bonanza riparian flood zone (Bonanza Riparian) in the Bonanza Creek Experimental Forest, Tanana Valley, Alaska. Box endpoints indicate the 25^th^ and 75^th^ percentile values; the black line and black point within the box indicate the median and mean value for each species, respectively. Data: Kielland et al. (2017). NA was used when sex was not provided."}
ggplot(data = juvenile_hares, aes(x = sex, y = weight, fill = sex)) +
  geom_beeswarm(aes(color = sex), width = 0.15, size = 1.5, alpha = 0.8) +
    scale_color_manual(values = c("navyblue", "indianred2")) +
    geom_boxplot(fill = NA, outlier.color = NA, width = 0.4) +
    stat_summary(fun = mean,
                 geom = 'point',
                 size = 2,
                 shape = 20) + # column graph
  facet_wrap(~ grid) + # facet wrap by grid for 
  labs(x = "Sex", y = "Average Weight (g)", title = "Figure 2: Juvenile Hare Weights by Sex and Location") +
  theme_minimal() + 
    theme(plot.title = element_text(face = "bold", hjust = 0.5), 
          legend.position = "None") # center the title and bold it
```

##	Male and Female Juvenile Hare Weight Comparisons

<center>**Table 1.** Summary Statistics of Juvenile Hare Trap Counts</center>
```{r}
juvenile_hares_summary <- juvenile_hares %>% 
  filter(sex %in% c("Male", "Female")) %>%
  drop_na(weight) %>% # filtering out NA sexes and weights
  group_by(sex) %>% # grouping by sex
 summarize( # finding summary statistics 
    mean_weight = round(mean(weight, na.rm = TRUE), 0), 
    median_weight = round(median(weight, na.rm = TRUE), 0), 
    sd_weight = round(sd(weight, na.rm = TRUE), 0), 
    sample_size = n())

juvenile_hares_summary %>% # creating a stylized table with the stats
  kable(caption = "Table 1 shows descriptive statistics (mean, standard deviation, sample size) of weight (g) of male and female juvenile snowshoe hares.", 
        col.names = c("Sex", 
                      "Mean Weight (g)", 
                      "Median Weight (g)",
                      "Standard Deviation (g)", 
                      "Sample Size")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```

```{r, include = FALSE}
# group by sex
juvenile_hares_sex <- juvenile_hares %>% 
  group_by(year) %>% 
  drop_na(sex)

# Exploratory graphs analyzing distribution
ggplot(data = juvenile_hares_sex, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~sex)
ggplot(data = juvenile_hares_sex, aes(x = weight)) +
  geom_histogram(bins = 15) +
  facet_wrap(~sex)
# data looks normally distributed for male and female for qq plots so I feel comfortable using a t-test
```

```{r, include = FALSE}
# storing vectors of male and female weights to use in t test

weight_male <- juvenile_hares %>% # storing male weights
  filter(sex == "Male") %>% 
  drop_na(weight) %>% 
  pull(weight)


weight_female <- juvenile_hares %>% # storing female weights
  filter(sex == "Female") %>% 
  drop_na(weight) %>% 
  pull(weight)

# running a t.test of the weights
weight_by_sex_t <- t.test(weight_male, weight_female)
weight_by_sex_t

# finding and storing the actual difference in means
sex_weight_mean_difference = mean(weight_by_sex_t$estimate[1]) - 
  mean(weight_by_sex_t$estimate[2])

# finding and storing standard deviations for results
male_sd <- sd(weight_male)
female_sd <- sd(weight_female)

# finding and storing the effect size
sex_weight_effect_size <- cohen.d(weight_male, weight_female, na.rm = TRUE)
sex_weight_effect_size
```

On average, male juvenile hares weighed `r round(juvenile_hares_summary$mean_weight[2], 1)` +/- `r round(juvenile_hares_summary$sd_weight[2], 1)` grams. On average, female juvenile hares weighed `r round(juvenile_hares_summary$mean_weight[1], 1)` +/- `r (juvenile_hares_summary$sd_weight[1])` grams. Male juvenile hares were found to weigh `r round(sex_weight_mean_difference, 1)` grams more than female juvenile hares on average.  QQplots and histograms were used to test assumptions of normalcy and homoscedasticity that allowed further statistical analyses. The results from the two sample t-test suggest there is a significant difference between the sample means of male and female juvenile hares (two-sample t-test, t = `r round(weight_by_sex_t$statistic, 3)`, p = `r round(weight_by_sex_t$p.value, 3)`), but the effect size was found to be small (Cohen's *d* = `r round(sex_weight_effect_size$estimate, 2)`), This means
that if samples of male and juvenile hares were drawn from populations with the same mean, the probability of taking two random samples with means at least as
different as the sample means found in this study by random chance is 0.07%. Therefore there is enough evidence to reject the null hypothesis. This analysis finds that the male and female juvenile hares in this sample were drawn from populations with the same mean.

## Juvenile Weight and Hindfoot Length Comparisons 
```{r, fig.cap = "Figure 3: Graph showing the hindfood length, in milimeters, of juvenile hares trapped between 1999 and 2012 in the Bonanza Creek Experimental Forest, Tanana Valley, Alaska. Data: Kielland et al. (2017). NA values were not included."}
ggplot(data = juvenile_hares_sex, aes(x = hindft, y = weight)) +
  geom_point(aes(color = sex), size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(x = "\nHindfoot Length (mm)",
       y = "Weight (g)\n",
       color = "Sex",
       title = "Figure 3: Juvenile Hare Hindfoot Length v Weight") +
  theme_minimal() + 
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
```
```{r}
juvenile_hare_cor <- cor.test(juvenile_hares$weight, juvenile_hares$hindft) # The Pearson Coefficient 
juvenile_hare_cor_tidy <- tidy(juvenile_hare_cor) # presentable
juvenile_hare_linear <- lm(weight ~ hindft, data = juvenile_hares) # Fitting a linear model
juvenile_hare_linear_glance <- glance(juvenile_hare_linear) # model-wide measures of uncertainty
juvenile_hare_linear_tidy <- tidy(juvenile_hare_linear) # Tidying the linear model
```

Across sex and location, the relationship between juvenile snowshoe hare is relatively linear. Based on our simple linear regression, we expect juvenile snowshoe hare weight to increase by `r round(juvenile_hare_linear$coefficient[2], 2)` g for every 1 mm increase in hindfoot length, on average (p < 0.001, R^2^ = `r round(juvenile_hare_linear_glance$r.squared, 2)`) with an average slope of $\beta$~1~ = `r round(juvenile_hare_linear_tidy$estimate[2], 2)` g mm^-1^. Hindfoot length and hare weight seem significantly positively correlated according to Pearson's Coefficient (Pearson's r = `r round(juvenile_hare_cor$estimate, 2)`, p < 0.001). However, Pearson's coefficient assumes that both samples are normally distributed and homoscedastic. Figure 3 shows heteroscedastic residuals for both the male and female juvenile hare weight. Additionally, exploratory plots omitted from this presentation showed an abnormal distribution in the weights of male juvenile hares. Therefore, a linear regression is not the most appropriate statistical analysis for the juvenile hare data provided in this study. 

# Summary

Exploratory data analysis reveals the following initial findings: 

- The number of juvenile hares trapped in the Bonanza Creek Experimental Forest in
Alaska has been decreasing over 13 years that this study was performed.  
- Male juveniles hares weighed significantly more than female juvenile hares.
- If samples of male and juvenile hares were drawn from populations with the same mean weight, the probability of taking two random samples with means at least as
different as the sample means found in this study by random chance is 0.07%.
- The data in this study would not be best represented by a linear regression analysis.

Next steps

- Compare the effect that location has on juvenile hare weight
- Determine a better statistical analysis to assess the relationship betwwen juvenile hare hindfood length and weight

# Citations

[Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative](https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14)
