---
title: "Exploratory Report on the weights of Juvenile Snowshoe Hares"
author: "Meghna Rao"
date: "11/7/2021"
output: html_document
code_folding: hide
---
## Introduction 
This report provides an exploratory overview of snowshoe hare
populations from 1998 - 2012 at the Bonanza Creek Experimental Forest in
Alaska (Kielland, K.et al. (2017). Specifically, the report evaluates
snowshoe hare trap counts over 13 years. It also compares juvenile hare
weights across males and females at three sites; Bonrip, Bonanza Mature,
and Bonanza Riparian. Finally, the report analyzes a potential
relationship between hind foot length (mm) and weights (g).

**Data Citation:** [Kielland, K., F.S. Chapin, R.W. Ruess, and Bonanza Creek LTER. 2017. Snowshoe hare physical data in Bonanza Creek Experimental Forest: 1999-Present ver 22. Environmental Data Initiative](https://doi.org/10.6073/pasta/03dce4856d79b91557d8e6ce2cbcdc14)

## Data and Analysis

Snowshoe hare trap count, weight and hind foot length data were
collected and made available by Senior Investigator Knut Kielland and
colleagues at the Bonanza Creek (LTER) site in Alaska. The data contain
observations for 378 juvenile hares, collected over 13 years (1998 -
2012). Following exploratory data visualization, juvenile hare male and
female weights are compared by two-sample t-tests using a significance
level (α) f 0.05 throughout. Differences between the groups are
described by Cohen's *d* effect size. The relationship between weight
and hind foot length is explored by a simple linear regression. All
analyses are in R version 4.0.2 using R studio version 1.2.1056.

## Juvenile hare initial investigation

```{r setup, include=FALSE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(here)
library(lubridate) # mutate date columns
library(kableExtra) # to show table statistics
library(effsize)
library(broom)
library(ggbeeswarm)
```
## Preliminary Analysis

```{r}
bonanza_hares <- read.csv(here("data", "bonanza_hares.csv"))

juvenile_hares <- bonanza_hares %>% 
  filter(age == "j") %>%   # Filter for just the juveniles
  mutate(date_new = mdy(date)) %>% # used lubridate library package to create a new column for the date
  mutate(year = year(date_new)) %>% 
   mutate(sex = case_when( # changing names of sex
    sex == "f" ~ "Female",
    sex == "m" ~ "Male",)) # created a new column of just the years
   
```

```{r}
juvenile_hares_count <- juvenile_hares %>% 
  count(year) # counted the number of juvenile bunnies per year

juvenile_hares_summary <- juvenile_hares_count %>% 
  summarize(mean_count = mean(n), # mean
            median_count = median(n), # median
            max_count = max(n), # max
            min_count = min(n)) # min
```


```{r, fig.cap= "Figure 1: Graph showing the total number of juvenile hares trapped between 1999 and 2012 in the Bonanza Creek Experimental Forest, Tanana Valley, Alaska. Data: Kielland et al. (2017)."}
ggplot(data = juvenile_hares_count, aes(x = year, y = n)) +
  geom_col(fill = "coral", width = 0.7) +
  labs(x = "Year", y = "Total Juvenile Hares Trapped", title = "Figure 1: Total Juvenile Hares Trapped by Year from 1998 to 2021 (RAO)") + # add labels to graph
  scale_y_continuous(breaks = c(0, 25, 50, 75, 100, 125)) + # manually putting tics for y axis
  scale_x_continuous(breaks = c(1998, 1999, 2000,2001, 2002, 2003, 2004, 2005, 
                                2006, 2007, 2008, 2009, 2010, 2011,2012)) + # manually putting ticks for x axis
  theme_minimal() +
    theme(plot.title = element_text(face = "bold", hjust = 0.5)) # center the title and bold it
```
### Annual Juvenile Hare Trap Counts. 

The average number of juvenile hares trapped per year in this study was `r round(juvenile_hares_summary$mean_count, 0)` and the median number of juvenile hares trapped per year was `r round(juvenile_hares_summary$median_count, 0)`. The greatest number of juvenile hares trapped was `r round(juvenile_hares_summary$max_count, 0)` in the first year of the study in 1999. There were no juvenile hares trapped in 2002 and 2009. The least number of juvenile hares trapped was `r round(juvenile_hares_summary$min_count, 0)` in 2012. To standardize the juvenile hare population in future analyses, the number of juvenile hare traps laid out every year, and the time of year the traps are laid out, needs to be consistent from year to year.

### Visualizing Juvenile Hare Weights by Sex and Site

We are showing summary statistics of 
```{r}
juvenile_hares_sex_site <- juvenile_hares %>% 
  group_by(sex, grid) %>% # group by sex, and grid
  summarize(mean_weight = mean(weight, na.rm = TRUE)) %>% # average weights by sex and grid, observations that were NA are included
  mutate(grid = case_when( # Rename sites in dataset so that the full name shows up in visualizations
    grid == "bonbs" ~ "Bonanza Black Spruce", 
    grid == "bonmat" ~ "Bonanza Mature",
    grid == "bonrip" ~ "Bonanza Riparian"))
```

```{r, fig.cap = "Figure 2: Graph showing summary statistics on the average weight of male and female juvenile hares trapped between 1999 and 2012 in mature lowland black spruce (Bonanza Black Spruce), mature black spruce forest (Bonanza Mature) and bonanza riparian flood zone (Bonanza Riparian) in the Bonanza Creek Experimental Forest, Tanana Valley, Alaska. Data: Kielland et al. (2017). NA was used when sex was not provided."}
ggplot(data = juvenile_hares_sex_site, aes(x = sex, y = mean_weight, fill = sex)) +
  geom_col() + # column graph
  facet_wrap(~ grid) + # facet wrap by grid for 
  labs(x = "Sex", y = "Average Weight (g)", title = "Figure 2: Juvenile Hare Weights by Sex and Location") +
  theme_minimal() + 
    theme(plot.title = element_text(face = "bold", hjust = 0.5), 
          legend.position = "None") # center the title and bold it
```
###	Juvenile weight comparison (male & female snowshoe hares).

<center>**Table 1.** Summary Statistics of Juvenile Hare Trap Counts</center>
```{r}
juvenile_hares_summary <- juvenile_hares %>% 
  filter(sex %in% c("Male", "Female")) %>%
  drop_na(weight) %>% # filtering out NA sexes and weights
  group_by(sex) %>% # grouping by sex
 summarize( # finding summary statistics 
    mean_weight = round(mean(weight, na.rm = TRUE), 0), 
    median_weight = round(median(weight, na.rm = TRUE), 0), 
    sd_weight = round(sd(weight, na.rm = TRUE), 0), 
    sample_size = n())

juvenile_hares_summary %>% # creating a stylized table with the stats
  kable(caption = "Table 1 shows descriptive statistics (mean, standard deviation, sample size) of weight (g) of male and female juvenile snowshoe hares.", 
        col.names = c("Sex", 
                      "Mean Weight (g)", 
                      "Median Weight (g)",
                      "Standard Deviation (g)", 
                      "Sample Size")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")
```
Further compare mean weights for **juvenile male and female snowshoe hares** included in this study. In this section, you should include:
A finalized table containing the mean, standard deviation, and sample size for male and female juvenile snowshoe hares, with a table caption (remember, a table caption should be placed *above* the table in your report)

```{r, include = FALSE}
# group by sex
juvenile_hares_sex <- juvenile_hares %>% 
  group_by(year)

# Exploratory graphs analyzing distribution
ggplot(data = juvenile_hares_sex, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~sex)
ggplot(data = juvenile_hares_sex, aes(x = weight)) +
  geom_histogram(bins = 15) +
  facet_wrap(~sex)
# data looks normally distributed for male and female for qq plots so I feel comfortable using a t-test
```

```{r, include = FALSE}
# storing vectors of male and female weights to use in t test

weight_male <- juvenile_hares %>% # storing male weights
  filter(sex == "Male") %>% 
  drop_na(weight) %>% 
  pull(weight)


weight_female <- juvenile_hares %>% # storing female weights
  filter(sex == "Female") %>% 
  drop_na(weight) %>% 
  pull(weight)

# running a t.test of the weights
weight_by_sex_t <- t.test(weight_male, weight_female)
weight_by_sex_t

# finding and storing the actual difference in means
sex_weight_mean_difference = mean(weight_by_sex_t$estimate[1]) - 
  mean(weight_by_sex_t$estimate[2])

# finding and storing standard deviations for results
male_sd <- sd(weight_male)
female_sd <- sd(weight_female)

# finding and storing the effect size
sex_weight_effect_size <- cohen.d(weight_male, weight_female, na.rm = TRUE)
sex_weight_effect_size
```

On average, male juvenile hares weighed `r round(juvenile_hares_summary$mean_weight[2], 1)` +/- `r round(juvenile_hares_summary$sd_weight[2], 1)` grams. On average, female juvenile hares weighed `r round(juvenile_hares_summary$mean_weight[1], 1)` +/- `r (juvenile_hares_summary$sd_weight[1])` grams. Male juvenile hares were found to weigh `r round(sex_weight_mean_difference, 1)` grams more than female juvenile hares on average.  GGplots and histograms were used to test assumptions of normalcy and homoscedasticity that allowed further statistical analyses. The null hypothesis for a Welch's two sample t-test assumes that the samples of male and female juvenile hares were pulled from populations with the same mean, meaning there would be no significant difference between the average weights of male and female juvenile hares in this study. The results from the two sample t-test suggest there is a significant difference between the sample means of male and female juvenile hares (two-sample t-test, t = `r round(weight_by_sex_t$statistic, 3)`, p = `r round(weight_by_sex_t$p.value, 3)`), but the effect size was found to be small (Cohen's *d* = `r round(sex_weight_effect_size$estimate, 2)`), This means
that if samples of male and juvenile hares were drawn from populations with the same mean, the probability of taking two random samples with means at least as
different as the sample means found in this study by random chance is 0.07%. Therefore there is enough evidence to reject the null hypothesis. This analysis finds that the male and female juvenile hares in this sample were drawn from populations with the same mean.

```{r, fig.cap = "Figure 3: Graph showing the hindfood length, in milimeters, of juvenile hares trapped between 1999 and 2012 in the Bonanza Creek Experimental Forest, Tanana Valley, Alaska. Data: Kielland et al. (2017). NA was used when sex was not provided."}
ggplot(data = juvenile_hares_sex, aes(x = hindft, y = weight)) +
  geom_point(aes(color = sex), size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(x = "\nHindfoot Length (mm)",
       y = "Weight (g)\n",
       color = "Sex",
       title = "Figure 3: Juvenile Hare Hindfoot Length v Weight") +
  theme_minimal() + 
    theme(plot.title = element_text(face = "bold", hjust = 0.5))
```
```{r}
# Pearson Coefficient and Linear Model
juvenile_hare_cor <- cor.test(juvenile_hares$weight, juvenile_hares$hindft) # The Pearson Coefficient 
juvenile_hare_cor_tidy <- tidy(juvenile_hare_cor) 
juvenile_hare_lm <- lm(weight ~ hindft, data = juvenile_hares) # Fitting a linear model
juvenile_hare_lm_glance <- glance(juvenile_hare_lm) 
juvenile_hare_lm_tidy <- tidy(juvenile_hare_lm) # Tidying the linear model
```

